Anomaly detection with kernel density estimation on manifolds

### 2022-05-22 paper/paper.Rmd
Remove spaces

The Riemannian metric using the method of @Perrault-Joncas2013-pq gives
some idea of the distortion of an embedding. Mapping the points through
a non-linear function "stretches" some regions of space and "shrinks"
others. The Riemannian gives us an idea of the direction and angle of
this stretching, which is informative for learning the manifold. In
variable kernel density estimate, the bandwidth matrix $H$ is also
defined to control the amount of smoothing for each data point.
Therefore, we propose to the use Riemannian metric as the bandwidth
matrix in, we could further get the kernel density estimation of the
manifold $M$. This kernel density estimate can then be used to produce
the highest density region plots [@Hyndman1996-lk] for outlier
visualization.

The rest of the paper is organized as follows. In \autoref{vkde}, we
present the proposed algorithm to detect anomalies based on variable
kernel density estimates of manifold embeddings. In this section, we
provide justification for the use of the Riemannian metric as the
bandwidth of variable kernel density estimation, including the
comparison with fixed bandwidth. \autoref{simulation} is composed of two
simulations with the proposed algorithm; the first deals with
2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data
and the second with a 5-D semi-hypersphere mapped in a 100-D space.
\autoref{application} contains the application to visualize and identify
anomalies in the Irish smart meter dataset. Conclusions and discussions
are presented in \autoref{conclusion}. Readers interested in the the
notions of Riemannian geometry mentioned in this paper could use the
Appendix in \autoref{riemgeo} as a reference.

@@

In general, given a multivariate random sample
$\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d$,
$i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel
density estimate of $f$ at point
$\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d$ is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
\end{equation} where $\pmb{H}_i$ is called the bandwidth matrix which is
a $d\times d$ symmetric positive definite matrix variable at each $i$,
$K(\pmb{y})$ is the kernel function, and
$K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})$.
When using the standard gaussian kernel
$K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})$, we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
\end{equation} with $|\pmb{H}_i|$ being the determinant of matrix
$\pmb{H}_i$.

@@

```{=tex}
\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $x_i \in \mathcal{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $y_i \in \mathcal{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters }
  \BlankLine
  \begin{algorithmic}[1]

@@

  \end{algorithmic}
\end{algorithm}
```
# Outlier detection on manifold learning embedding

@@

```{=tex}
\begin{algorithm}[!htb]
  \caption{Variable kernel density estimates with Riemannian metric}
  \label{alg:vkderm}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
  \BlankLine
  \begin{algorithmic}[1]

  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;

  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;

  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \@ref(equ:denestimator);

  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.

  \end{algorithmic}
\end{algorithm}
```
If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could
be used to identify the relative location of outliers in the embedding
and find which observations lie in the highest density region of
specified coverage, eg. 1%, 50%, 99%, \>99%.

@@

We first generate a 2-dimensional data of size $N=2000$ from a Gaussian
mixture model with four components with different means
$\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$
and the same variance-covariance matrix
$\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$. The mixture proportions
are equally set as $\pi_i=0.25, i=1,2,3,4$.
<!-- $\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)$. -->
Then the mixture Gaussian mixture density function is a weighted linear
combination of the four component Gaussian densities as \begin{equation}
\label{eq:gmm}
P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
\end{equation} \autoref{fig:metadensity} shows the 2-dimensional meta
data and the colors indicate the true density of all data points
calculated from \@ref(eq:gmm), with brighter colors showing higher
densities and darker colors showing lower densities. We then define
outliers as points with lowest densities shown in black and typical
points with highest densities shown in yellow. Based on the true density
plot, the outliers are scattered in the middle and the outer area of the
whole structure, while typical points are near the means of four
kernels. These are *true outliers* to be compared with outliers from the
kernel density estimates.

@@

Now we are able to apply different manifold learning algorithms to
$(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to
$d=2$, and further estimate the density of the 2-D embedding. According
to the density estimates, we could rank the data points and then
identify which observations lie in the highest density region of
specified coverage, eg. 1%, 5%, 50%, 99%, \>99%. For each of the five
manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps,
t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding
plot in the same row, with the colors indicating the densities levels,
the left column for true densities from the Gaussian mixture model, the
middle column for highest density region plots with densities from our
proposed variable KDE method, and the right for similar HDR plots with
densities from KDE with fixed bandwidth. The top twenty outliers with
the lowest densities are highlighted in black with point indexes in
blue. From \autoref{fig:metadensity} and the data generating process, we
know that there are four highest density regions. However, in all
manifold learning embeddings colored with true densities (left column in
\autoref{fig:sroutliers} ), except for LLE, the number of highest
density regions are not the same as the meta data. When comparing the
number of HDRs for variable and fixed bandwidth (middle and right column
in \autoref{fig:sroutliers} ), our proposed method with variable
bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian
Eigenmaps (top three rows in \autoref{fig:sroutliers} ). In terms of the
top 20 outliers found rowwise, variable bandwidth could find most
outliers lying on the left area of the embedding in ISOMAP and UMAP, and
both methods in LLE embedding could find the outliers in the outer area,
but for the other methods, both variable and fixed bandwidth are not
detected true outliers accurately. For t-SNE and UMAP embedding, the
embedding structure is highly distorted and the points are clustered
together in a discontinuous way, which is also shown in the clustered
outliers.

@@

Then we initial the other 95 dimensions in the high-dimensional space as
zero columns and further rotate the $100$-dimensional data of size $N$
(denote the transpose of the data matrix as $\pmb{X}_0$ with dimension
$100 \times N$) to get rid of the zeros so that it could be passed to
the manifold learning algorithms. The rotation matrix is derived from
the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all
components randomly generated from a uniform distribution
$\mathcal{U}(0,1)$. For any real matrix $\pmb{A}$ of dimension
$p\times q$, the QR decomposition could decompose the matrix into the
multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that
$\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit
norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and
$\pmb{R}$ is an upper triangular matrix. Matrix $\pmb{Q}$ satisfies
$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that
the pairwise Euclidean distances between data points in
$\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
Therefore, we use matrix $Q$ as the rotation matrix for where the
rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension
$N \times 100$ is now the input data for the manifold learning
algorithms. Again, we set the embedding dimension to be equal to the
meta data dimension $d=5$.

@@

In this paper, we propose a new method to estimate the density of
manifold learning embedding and further identify outliers based on the
densities. The Riemannian metric measure the direction and angle of the
distortion when mapping data points through a non-linear function in
manifold learning algorithms. By introducing the Riemannian metric as
the pointwise variable bandwidth matrix in kernel density estimation,
the local distortion in the low-dimensional embedding could be used to
estimate densities, leading to a more accurate description of the data
distributions. We compare our proposed method with fixed bandwidth KDE
by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or
twin peaks data and 5-D semi-hypersphere meta data mapped in 100-D
space, and show that variable bandwidth could improve the density
estimation given a good manifold learning embedding.

@@

```{=html}
<!-- The aim of the project is not to evaluate different manifold learning algorithms.  Although ISOMAP worked OK here it is still distorted and in real examples, you will never 'know' that the data are uniform on the sphere.  Instead, the idea of this project is to do kernel density estimation in a way that takes distortion into account.
 -->
```
# Acknowledgment {.unnumbered}

>>>>

The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding. Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others. The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold. In variable kernel density estimate, the bandwidth matrix $H$ is also defined to control the amount of smoothing for each data point.
Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold $M$. This kernel density estimate can then be used to produce the highest density region plots[@Hyndman1996-lk] for outlier visualization.

The rest of the paper is organized as follows. In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings. In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth. \autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset. Conclusions and discussions are presented in \autoref{conclusion}. Readers interested in the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.

@@

In general, given a multivariate random sample $\pmb{Y}_i=(Y_{i1}, Y_{i2}, \dots, Y_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$ drawn from a density $f$, the variable kernel density estimate of $f$ at point $\pmb{y}=(y_1, y_2, \dots, y_d)^\prime \in \mathcal{R}^d$ is given by
\begin{equation}
\label{eq:kde}
\hat{f}(\pmb{y})=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i),
\end{equation} 
where $\pmb{H}_i$ is called the bandwidth matrix which is a $d\times d$ symmetric positive definite matrix variable at each $i$, $K(\pmb{y})$ is the kernel function, and $K_{\pmb{H}_i}(\pmb{y}) = |\pmb{H_i}|^{-1/2} K(\pmb{H_i}^{-1/2}\pmb{y})$.
When using the standard gaussian kernel, $K(\pmb{y}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{y}'\pmb{y})$, we have
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{y}-\pmb{Y}_i)=(2\pi)^{-d/2}|\pmb{H_i}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{Y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{Y}_i)\right],
\end{equation} 
with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.

@@

<!-- ```{=tex} -->
\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $x_i \in \mathcal{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $y_i \in \mathcal{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters }
  \BlankLine
  \begin{algorithmic}[1]

@@

  \end{algorithmic}
\end{algorithm}

# Outlier detection on manifold learning embedding

@@

\begin{algorithm}[!htb]
  \caption{Variable kernel density estimates with Riemannian metric}
  \label{alg:vkderm}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
  \BlankLine
  \begin{algorithmic}[1]

  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $\pmb{g}_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;
  \STATE Set the variable bandwidth for each observation as $\pmb{h}_i = \pmb{g}_i$;
  \STATE Compute the kernel density estimates $f_N(\pmb{y}_i)$ for all $i=1,\ldots,N$ using Equation \@ref(equ:denestimator);
  \STATE Reorder the embedding coordinates $\pmb{y}$ according to the density estimates $f_N(y)$ and subset the top \textit{n\_outliers} as the outliers.
  \end{algorithmic}
\end{algorithm}

If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg. 1%, 50%, 99%, \>99%.

@@

We first generate a 2-dimensional data of size $N=2000$ from a Gaussian mixture model with four components with different means
$\pmb{\mu_1}=(0.25, 0.25)^\prime, \pmb{\mu_2}=(0.25, 0.75)^\prime, \pmb{\mu_3}=(0.75, 0.25)^\prime, \pmb{\mu_4}=(0.75, 0.75)^\prime$ and the same variance-covariance matrix $\pmb{\Sigma}_i=diag(0.02, 0.02), i=1,2,3,4$. The mixture proportions
are equally set as $\pi_i=0.25, i=1,2,3,4$.
<!-- $\big(\begin{smallmatrix} 0.02&0\\ 0&0.02 \end{smallmatrix}\big)$. -->
Then the mixture Gaussian mixture density function is a weighted linear combination of the four component Gaussian densities as 
\begin{equation}
\label{eq:gmm}
P(\pmb{X}=\pmb{x}) = \sum_{i=1}^{4}\pi_i \frac{1}{(2\pi)^{d/2}|\pmb{\Sigma}_i|^{-1/2}} \exp{\{-\frac{1}{2} (\pmb{x}-\pmb{\mu_i})^\prime \pmb{\Sigma}_i^{-1} (\pmb{x}-\pmb{\mu_i}) \}}.
\end{equation}

\autoref{fig:metadensity} shows the 2-dimensional meta data and the colors indicate the true density of all data points calculated from \@ref(eq:gmm), with brighter colors showing higher densities and darker colors showing lower densities. We then define outliers as points with lowest densities shown in black and typical points with highest densities shown in yellow. Based on the true density plot, the outliers are scattered in the middle and the outer area of the whole structure, while typical points are near the means of four kernels. These are *true outliers* to be compared with outliers from the
kernel density estimates.

@@

Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding. According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg. 1%, 5%, 50%, 99%, \>99%. For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels,
the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth. The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue. From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions. However, in all
manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data. When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ). In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detecting true outliers accurately. For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.

@@

Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the $100$-dimensional data of size $N$(denote the transpose of the data matrix as $\pmb{X}_0$ with dimension $100 \times N$) to get rid of the zeros so that it could be passed to the manifold learning algorithms. The rotation matrix is derived from the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$. For any real matrix $\pmb{A}$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that $\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and $\pmb{R}$ is an upper triangular matrix. Matrix $\pmb{Q}$ satisfies
$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the pairwise Euclidean distances between data points in
$\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$.
Therefore, we use matrix $Q$ as the rotation matrix for where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the manifold learning algorithms. Again, we set the embedding dimension to be equal to the meta data dimension $d=5$.

@@

In this paper, we propose a new method to estimate the density of manifold learning embedding and further identify outliers based on the densities. The Riemannian metric measure the direction and angle of the distortion when mapping data points through a non-linear function in manifold learning algorithms. By introducing the Riemannian metric as the pointwise variable bandwidth matrix in kernel density estimation, the local distortion in the low-dimensional embedding could be used to estimate densities, leading to a more accurate description of the data distributions. We compare our proposed method with fixed bandwidth KDE by two simulation settings, 2-D meta data mapped as a 3-D swiss roll or twin peaks data and 5-D semi-hypersphere mapped in 100-D space, and show that variable bandwidth could improve the density estimation given a good manifold learning embedding.

@@

<!-- The aim of the project is not to evaluate different manifold learning algorithms.  Although ISOMAP worked OK here it is still distorted and in real examples, you will never 'know' that the data are uniform on the sphere.  Instead, the idea of this project is to do kernel density estimation in a way that takes distortion into account.
-->

# Acknowledgment {.unnumbered}
